A Toolbox for Manipulating and Assessing Color Palettes

### 2019-08-26 pkg/colorspace/paper/Toolbox/article.Rmd
added discussion of human perceptual axes and corresponding implementation in HCL
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@501 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

Various utilities are available for working with the implemented \proglang{S}4 color spaces:

@@

>>>>

## Human color vision and the HCL color model

It has been hypothesized that human color vision has evolved in three distinct stages:

1. Perception of light/dark contrasts (monochrome only).
2. Yellow/blue contrasts (usually associated with our notion of warm/cold colors).
3. Green/red contrasts (helpful for assessing the ripeness of fruit).

See @color:Kaiser+Boynton:1996, @color:Knoblauch:2002, @color:Ihaka:2003, @color:Lumley:2006, @color:Zeileis+Hornik+Murrell:2009
for more details and references. Thus, colors can be described using a 3-dimensional space as shown in the left panel
of Figure\ \ref{fig:hcl-axes}.
However, for describing colors in such a space, it is more natural for humans to employ
polar coordinates in the color plane (yellow/blue vs.\ green/red, visualized by the dashed circle in Figure\ \ref{fig:hcl-axes})
plus a third light/dark axis.
Hence, color models that attempt to capture these perceptual axes are also called perceptually-based color spaces.
As already argued above, the HCL model captures these dimensions very well, calling them: _hue_, _chroma_, and _luminance_.
The corresponding sRGB gamut, i.e., the HCL colors that can also be represented in sRGB, is visualized
in the right panel of Figure\ \ref{fig:hcl-axes} [by @color:Horvath+Lipka:2016]. An animated version of the same plot
is provided online by @color:Horvath+Lipka:2017.

```{r human-axes, echo = FALSE, fig.width = 4.5, fig.height = 4.5, fig.show = "hide"}
par(mar = rep(0, 4))
midgray <- gray(134/255)
res <- persp(c(-1, 1), c(-1, 1), matrix(c(-1, 1, 1, -1), 2, 2),
  xlim = c(-0.77, 0.77), ylim = c(-0.77, 0.77), zlim = c(-0.77, 0.77),
  border = "transparent", box = FALSE, axes = FALSE, phi = 20, theta = 30, r = 50)
ax <- trans3d(c(1, -1, 0, 0, 0, 0), c(0, 0, 1, -1, 0, 0), c(0, 0, 0, 0, 1, -1), res)
circ <- (-45:45)/50
circ <- trans3d(rep(circ, 2), c(sqrt(circ[1]^2 - circ^2), -sqrt(circ[1]^2 - circ^2)), 0, res)
arrows(ax$x[c(1, 3, 5)], ax$y[c(1, 3, 5)], ax$x[c(2, 4, 6)], ax$y[c(2, 4, 6)], code = 3, length = 0.1, col = midgray, lwd = 3)
lines(circ$x, circ$y, col = midgray, lty = 2)
text(ax$x, ax$y, c("red", "green", "yellow", "blue", "light", "dark"), col = midgray, pos = c(4, 2, 4, 2, 3, 1))
```

\begin{figure}[t!]
\centering 
\includegraphics[width=0.42\textwidth]{Figures/human-axes-1.pdf} 
\includegraphics[width=0.49\textwidth]{Figures/hcl-wikimedia.png}
\caption{Visualization of axes capturing human color vision (left) and the corresponding HCL color model (right).}\label{fig:hcl-axes}
\end{figure}

The shape of the HCL space is a distorted double cone which is seen best by looking at
vertical slices, i.e., chroma-luminance planes for given hues. For example, the left panel
in Figure\ \ref{fig:hcl-projections} depicts
the chroma-luminance plane for a certain blue (hue = 255). Along with luminance the colors change from
dark to light. With increasing chroma the colors become more colorful, where the highest chroma is
possible for intermediate luminance.

As some colors are relatively dark (e.g., blue and red assume their maximum chroma for relatively low luminances)
while others are relatively light (e.g., yellow and green), horizontal slices of hue-chroma planes for given hue
have somewhat irregular shapes. The right panel in Figure\ \ref{fig:hcl-projections} shows such a
hue-chroma plane for moderately light colors (luminance = 70).
At that luminance, green and orange can become much more colorful compared to blue or red.

```{r hcl-projections, echo = FALSE, fig.width = 9.5, fig.height = 4.5, dev = "png", fig.cap = "Vertical (left) and horizontal (right) slices of the HCL space yielding a chroma-luminance plane for given hue and a hue-chroma plane for given luminance, respectively."}
par(mfrow = c(1, 2))
hclplot(sequential_hcl(5), h = 255, c = 125, size = 0, lwd = 0)
hclplot(qualitative_hcl(5), l = 70, c = 125, size = 0, lwd = 0)
```

@@

Several utilities are available for working with the \proglang{S}4 classes implementing the
color spaces listed above.

@@

Meanwhile, base \proglang{R} has also adopted the HCL-based palettes from \pkg{colorspace} into the `hcl.colors()`
function in \pkg{grDevices} [@color:grDevices]. This provides all the named palettes introduced in
\pkg{colorspace} (with the same names, and defaulting to `"Viridis"`) but without the flexibility
to modify or adapt existing palettes.
